Automated inference of production rules for glycans

### 2021-03-29 papers/sugar-synth/main.tex
variations first cut
a few issues fixed

\input{macro.tex}

\title{%Automated inference
  Synthesis of production rules for glycans}
\author{Ansuman Biswas, Ashutosh Gupta, Meghana Missula, and Mukund Thattai}
%\institute{IITB and NCBS}

@@

\begin{abstract}
  Glycans are tree-like polymers made up of sugar monomer building blocks. They are found on the surface of all living cells, and distinct glycan trees act as identity markers for distinct cell types. Proteins called GTase enzymes assemble glycans via the successive addition of monomer building blocks. The rules by which the enzymes operate are not fully understood. In this paper, we present a novel SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans from a cell. We have build a tool based on the method and applied it to infer rules based on published glycan data.
\end{abstract}

@@

\section{Related work}
\label{sec:related}
\input{related.tex}

>>>>

\input{macro.tex}
\setlength{\textfloatsep}{0pt}%

\title{Automated inference of production rules for glycans}
% \author{Ansuman Biswas, Ashutosh Gupta, Meghana Missula, and Mukund Thattai}
%\institute{IITB and NCBS}

@@

\begin{abstract}
  Glycans are tree-like polymers made up of sugar monomer building blocks. They are found on the surface of all living cells, and distinct glycan trees act as identity markers for distinct cell types. Proteins called GTase enzymes assemble glycans via the successive addition of monomer building blocks. The rules by which the enzymes operate are not fully understood. In this paper, we present a novel SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans from a cell. We have built a tool based on the method and applied it to infer rules based on published glycan data.
\end{abstract}

@@

% \section{Related work}
% \label{sec:related}
% \input{related.tex}

### 2021-03-29 papers/sugar-synth/variations.tex

{\em Compartments : } The production rules may live in different ordered compartments.
% The compartments are ordered.
A molecule moves from one compartment to the next.
The rules of the current compartment apply to the molecule.
% We have presented the formal description of the compartments in
% section~\ref{sec:model}.
To support the compartments, we take an additional integer
input $k$ in $\textsc{SugarSynth}$ to indicate the maximum number
of compartments.
We construct each template rule with a new variable with domain $[1,k]$.
Let $v_r$ be a node of some template rule. We write $compart(v_r)$
for the compartment variable for the template.

@@

{\em Unbounded molecules: }
We only observe a finite set of molecules in cells.
However, a set of production rules may be capable of producing an unboundedly large number of molecules.
In such cases, rules produce molecules the have repeating patterns of a subtree while rest of the tree being exactly same as one of the input molecules. The rules may be acceptable in some biological settings. We modify constraints to not declare such molecules as negative.
Let us suppose we have a repeat pattern of depth $d$ with $r$ repetitions.
Let $v$ be the node in template molecule where repetition has begin.
We define $\textsc{RepeatHeads}(v,r,d)$ that returns nodes $v_0,.... v_r$ such that
$v_1 = v$, $v_i$ is the $d$th ancestor of $v_{i+1}$.
% Those rules should in the same compartment and are slow.
Let us define \textsc{Repeat} and \textsc{Exact}, which encodes that the trees rooted at $v_{i}$s repeat.\\
$\textsc{Repeat}([v_0,...,v_r], v') :=  \textsc{Exact}(v',v_r,\bot) \land \Land_{i\in[0,r-1]} \textsc{Exact}( v_i, v_{i+1},v_{i+1})$\\
% $\textsc{Exact}(v,v') := (M(v) = M(v')) \land \Land_{i \in [1,w]} \textsc{Exact}(C(v,i),C(v',i))$ \\
% $\textsc{Exact}(\bot,v') := \bot, \textsc{Exact}(v,\bot) := \bot, \textsc{Exact}(\bot,\bot) := \bot$\\
$\textsc{Exact}(\bot,v',\_) := \lfalse, \textsc{Exact}(v,\bot,\_) := \lfalse,$\\
$ \textsc{Exact}(\bot,\bot,\_) := \ltrue,\textsc{Exact}(v_s,\_,v_s) := \ltrue$\\
$\textsc{Exact}(v,v',v_s) := (M(v) = M(v')) \land \Land_{i \in [1,w]} \textsc{Exact}(C(v,i),C(v',i),v_s)$\\
% $\textsc{Repeat}(v, v', r, d) :=  \textsc{Repeat}(v, ancestor(v'), r, d-1)$\\
% $\textsc{Repeat}(v, v', r, 0) := (compart(v) = compart(v')) \land \textsc{Repeat}(v, C^{2d}(v'), r-1, d)$ \\
% $\textsc{Repeat}(v, v', 0, d) :=  \textsc{ExactMatch}(v,C^{2d}(v'))$ \\
% $\textsc{Repeat}(\bot, \bot, r, d) :=  \top  \quad \textsc{Repeat}(v, \bot, r, d) :=  \bot
% \quad \textsc{Repeat}(\bot, v', r, d) :=  \bot $\\
% \begin{align*}
%   \textsc{Repeat}(m, m', r, d) :=\;&  \textsc{Repeat}(m, ancestor(m'), r, d-1)\\
%   \textsc{Repeat}(m, m', r, 0) :=\;& (compart(m) = compart(m')) \land \textsc{Repeat}(m, C^{2d}(m'), r-1, d) \\
%   \textsc{Repeat}(m, m', 0, d) :=\;&  \textsc{ExactMatch}(m,C^{2d}(m')) \\
%   \textsc{Repeat}(\bot, \bot, r, d) :=\;&  \top \\
%   \textsc{Repeat}(m, \bot, r, d) :=\;&  \bot \\
%   \textsc{Repeat}(\bot, m', r, d) :=\;&  \bot \\
%   \textsc{ExactMatch}(m,m') :=\;& (M(m) = M(m')) \land \Land_{i \in [1,w]} \textsc{ExactMatch}(C(m,i),C(m',i)) \\
%   \textsc{ExactMatch}(\bot,m') :=\;& \bot \\
%   \textsc{ExactMatch}(m,\bot) :=\;& \bot 
% \end{align*}
  % where $ancestor(m')$ is the parent node of $m'$ and $C^{2*d}$ is the application of $C(m',i)$, $2d$ times and $i$ comes from the reverse of the path traversed by m' to reach it's $d^{th}$ parent.
%   The constraints for allowing runaway reactions are added after~\ref{line:consNewR} to modify \textsc{SugarSynth}.\\
% \begin{algorithmic}[1]
%   \vspace{1ex}
%   \setcounterref{ALG@line}{line:consNewR}
%   \State $mCons \landplus \Lor_{1 \leq r \leq r_0} \Lor_{1 \leq d \leq d_0} \Lor_{m \in \mu} (M(m) \neq M(\hat{m})) \implies  \textsc{Repeat}(m, \hat{m}, r, d) $
%   \vspace{1ex}
% \end{algorithmic}
% where \textsc{Repeat} adds the constraints described above due to which runaway reactions are allowed.
We modify the third constraints of $ \textsc{MolTemplateCorrectness}(\hat{m},\mu)$, which encodes that negative molecules not in $\mu$. We change the definition of $Neq$ as follows.\\
$Neq(v, v') :=\; (\nu(v) \neq M(v') \land \lor_{r\in[1,r_0],d\in[1,d_0]}\textsc{Repeat}(\textsc{RepeatHeads}(v,r,d), v')) $\\
\mbox{}\hspace{30mm}$\lor \Lor_{{i \in [1,w]}} Neq( C(v,i), C(v',i ) ),$\\
where $d_0$ and $r_0$ are limits on the depth of the repeating subtrees and the number of repetitions respectively.
The change will accept molecules with repeating patterns as positive samples.

\paragraph{Non-monotonic rules :}
Some production rules can not be applied if another node is present in a sibling.
We call such rule non-monotonic because it may get disabled as the molecule grows.
This feature of rules help in producing an exact set of desired molecules.
We add an extra bit on each node of rule
template called $\textsc{HardEnds}$.
If the node is absent, its parent is present, and  $\textsc{HardEnds}$ bit is true, then
no node must be present in the matching pattern at the time of the application of the rule.
We modify the function $\textsc{MatchTree}$ by inserting the following constraints after~\ref{line:matchtree-cons}:
\begin{minipage}{1.0\linewidth}
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:matchtree-cons}
  \State $c \landplus \textsc{HardEnds}(v_r) \implies (mark \leq \tau(v))$
\end{algorithmic}  
\end{minipage}
The constraints state that if the applicable rule has \textsc{HardEnds}, then it has to be added at a time earlier than the current time of the molecule, effectively restricting the addition of further rules.

>>>>

{\em Compartments : } The production rules may live in different compartments.
% The compartments are ordered.
A molecule moves from one compartment to the next.
% The rules of the current compartment apply to the molecule.
% We have presented the formal description of the compartments in
% section~\ref{sec:model}.
To support the compartments, we take an additional integer
input $k$ in $\textsc{SugarSynth}$ to indicate the maximum number
of compartments.
We construct each template rule with a new variable with domain $[1,k]$.
Let $v_r$ be a node of some template rule. We write $compart(v_r)$
for the compartment variable for the template.

@@

{\em Unbounded molecules: }
We only observe a finite set of molecules in cells.
However, a set of production rules may be capable of producing an unboundedly large number of molecules.
In such cases, rules produce molecules the have repeating patterns of a subtree while rest of the tree being exactly same as one of the input molecules. The rules may be acceptable in some biological settings. We modify constraints to not declare such molecules as negative.
Let us suppose we have a repeat pattern of depth $d$ with $r$ repetitions.
Let $v$ be the node in template molecule where repetition has begin.
We define $\textsc{RepeatHeads}(v,r,d)$ that returns nodes $v_0,.... v_r$ such that
$v_1 = v$, $v_i$ is the $d$th ancestor of $v_{i+1}$.
% Those rules should in the same compartment and are slow.
Let us define \textsc{Repeat} and \textsc{Exact}, which encodes that the trees rooted at $v_{i}$s repeat.\\
$\textsc{Repeat}([v_0,...,v_r], v') :=  \textsc{Exact}(v',v_r,\bot) \land \Land_{i\in[0,r-1]} \textsc{Exact}( v_i, v_{i+1},v_{i+1})$\\
% $\textsc{Exact}(v,v') := (M(v) = M(v')) \land \Land_{i \in [1,w]} \textsc{Exact}(C(v,i),C(v',i))$ \\
% $\textsc{Exact}(\bot,v') := \bot, \textsc{Exact}(v,\bot) := \bot, \textsc{Exact}(\bot,\bot) := \bot$\\
$\textsc{Exact}(\bot,v',\_) := \lfalse, \textsc{Exact}(v,\bot,\_) := \lfalse,$\\
$ \textsc{Exact}(\bot,\bot,\_) := \ltrue,\textsc{Exact}(v_s,\_,v_s) := \ltrue$\\
$\textsc{Exact}(v,v',v_s) := (M(v) = M(v')) \land \Land_{i \in [1,w]} \textsc{Exact}(C(v,i),C(v',i),v_s)$\\
% $\textsc{Repeat}(v, v', r, d) :=  \textsc{Repeat}(v, ancestor(v'), r, d-1)$\\
% $\textsc{Repeat}(v, v', r, 0) := (compart(v) = compart(v')) \land \textsc{Repeat}(v, C^{2d}(v'), r-1, d)$ \\
% $\textsc{Repeat}(v, v', 0, d) :=  \textsc{ExactMatch}(v,C^{2d}(v'))$ \\
% $\textsc{Repeat}(\bot, \bot, r, d) :=  \top  \quad \textsc{Repeat}(v, \bot, r, d) :=  \bot
% \quad \textsc{Repeat}(\bot, v', r, d) :=  \bot $\\
% \begin{align*}
%   \textsc{Repeat}(m, m', r, d) :=\;&  \textsc{Repeat}(m, ancestor(m'), r, d-1)\\
%   \textsc{Repeat}(m, m', r, 0) :=\;& (compart(m) = compart(m')) \land \textsc{Repeat}(m, C^{2d}(m'), r-1, d) \\
%   \textsc{Repeat}(m, m', 0, d) :=\;&  \textsc{ExactMatch}(m,C^{2d}(m')) \\
%   \textsc{Repeat}(\bot, \bot, r, d) :=\;&  \top \\
%   \textsc{Repeat}(m, \bot, r, d) :=\;&  \bot \\
%   \textsc{Repeat}(\bot, m', r, d) :=\;&  \bot \\
%   \textsc{ExactMatch}(m,m') :=\;& (M(m) = M(m')) \land \Land_{i \in [1,w]} \textsc{ExactMatch}(C(m,i),C(m',i)) \\
%   \textsc{ExactMatch}(\bot,m') :=\;& \bot \\
%   \textsc{ExactMatch}(m,\bot) :=\;& \bot 
% \end{align*}
  % where $ancestor(m')$ is the parent node of $m'$ and $C^{2*d}$ is the application of $C(m',i)$, $2d$ times and $i$ comes from the reverse of the path traversed by m' to reach it's $d^{th}$ parent.
%   The constraints for allowing runaway reactions are added after~\ref{line:consNewR} to modify \textsc{SugarSynth}.\\
% \begin{algorithmic}[1]
%   \vspace{1ex}
%   \setcounterref{ALG@line}{line:consNewR}
%   \State $mCons \landplus \Lor_{1 \leq r \leq r_0} \Lor_{1 \leq d \leq d_0} \Lor_{m \in \mu} (M(m) \neq M(\hat{m})) \implies  \textsc{Repeat}(m, \hat{m}, r, d) $
%   \vspace{1ex}
% \end{algorithmic}
% where \textsc{Repeat} adds the constraints described above due to which runaway reactions are allowed.
We modify constraints of $ \textsc{MolTemplateCorrectness}(\hat{m},\mu)$, which encodes that negative molecules are not in $\mu$. We replace the definition of $Neq$ as follows.\\
$Neq(v, v') :=\; (\nu(v) \neq M(v') \land \lor_{r\in[1,r_0],d\in[1,d_0]}\textsc{Repeat}(\textsc{RepeatHeads}(v,r,d), v')) $\\
\mbox{}\hspace{30mm}$\lor \Lor_{{i \in [1,w]}} Neq( C(v,i), C(v',i ) ),$\\
where $d_0$ and $r_0$ are limits on the depth of the repeating subtrees and the number of repetitions respectively.
The change will accept molecules with repeating patterns as positive samples.

\paragraph{Non-monotonic rules :}
Some production rules are not applicable if another node is present in a sibling.
We call such a rule non-monotonic because it may get disabled as the molecule grows.
This feature of rules helps in producing an exact set of desired molecules.
We add an extra bit on each node of rule
template called $\textsc{HardEnds}$.
If the node is absent, its parent is present, and  $\textsc{HardEnds}$ bit is true, then
no node must be present in the matching pattern at the time of the application of the rule.
We modify the function $\textsc{MatchTree}$ by inserting the following constraints after~\ref{line:matchtree-cons}:
\begin{minipage}{1.0\linewidth}
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:matchtree-cons}
  \State $c \landplus \textsc{HardEnds}(v_r) \implies (mark \leq \tau(v))$
\end{algorithmic}  
\end{minipage}
The constraints state that if the applicable rule has \textsc{HardEnds}, then it has to be added at a time earlier than the current time of the molecule, effectively restricting the addition of further rules.
